The Impact of Alternation

### 2017-05-02 draft.tex
ok

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$. We denote by
$\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$ and write
$\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
occurs only within literals and in disjunctive normal form (DNF) if it
is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
\bigwedge_{j=1}^{M_i} \ell_{ij}$.

@@

An \emph{alternating data automaton} (or, simply, and automaton) is a
tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, also called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
  \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final} states, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
current values of the variables of $\A$. Formally, let $\vec{x}_k =
\set{x_k \mid x \in \vec{x}}$, $k\geq0$, be a time-stamped set of
variables. For an input even $a \in \Sigma$ and a formula $\phi$ we
write $\Delta^k(\phi,a)$ for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$.

@@

For a formula $\phi \in \Form^+(Q,\vec{x})$, with no negated
occurrences of the boolean variables in $Q$, we define the
formula $\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on
the structure of $\phi$:
\[\begin{array}{lclclclc}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\
\overline{\phi} & \equiv & \phi & \text{ if $\phi \in Q$}\hspace*{1cm} &
\overline{\phi} & \equiv & \neg\phi & \text{if $\phi \not\in Q$ is an atom}
\end{array}\]
If $\phi$ does not contain variables from $Q$, then $\overline{\phi}$
and $\neg\phi$ are equivalent, and in general:

@@

@@

We thus turn to semi-algorithms for this problem and develop methods
that are supposed to be both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

>>>>

We consider an infinite countable set of \emph{variables} $\vars$,
where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
between two terms of sort $\Data$. A \emph{literal} is an atom or its
negation. A \emph{formula} is a combination of atoms using disjunction
$\vee$, conjunction $\wedge$ and negation $\neg$. We denote by
$\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$ and write
$\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
replacing each occurrence of $x$ by $t$. For indexed sets
$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
occurs only within literals and in disjunctive normal form (DNF) if it
is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
\bigwedge_{j=1}^{M_i} \ell_{ij}$. The size of a formula $\phi$,
denoted $\len{\phi}$ is the number of symbols occuring in it.

@@

An \emph{alternating data automaton} (or, simply, and automaton) is a
tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
%
\item $\vec{x} \subset \vars$ is a finite set of variables of sort
  $\Data$,
%
\item $Q$ is a finite set of variables of sort $\Bool$, also called
  \emph{states},
%
\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
  \emph{initial configurations},
%
\item $F \subseteq Q$ is a set of \emph{final} states, and
%
\item $\Delta : Q \times \Sigma \rightarrow
  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
  \mid x \in \vec{x}\}$.
\end{compactitem}
Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
current values of the variables of $\A$. Formally, let $\vec{x}_k =
\set{x_k \mid x \in \vec{x}}$, $k\geq0$, be a time-stamped set of
variables. For an input even $a \in \Sigma$ and a formula $\phi$ we
write $\Delta^k(\phi,a)$ for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. The size of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclclc}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\
\overline{\phi} & \equiv & \phi & \text{ if $\phi \in Q$}\hspace*{1cm} &
\overline{\phi} & \equiv & \neg\phi & \text{if $\phi \not\in Q$ is an atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
equivalent, and in general:

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be
implemented in linear time. This is not very surprising, since linear
time intersection and complementation are possible for alternating
automata over finite alphabets \cite{ChandraKozenStockmeyer81}.

@@

Consequently, we give up on the guarantee for termination and develop
semi-algorithmic methods that are supposed to be
both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
finite input alphabet is $\Sigma$, for the rest of this section. Given
a finite sequence $u = a_1\ldots a_n \in \Sigma^*$ of input events and
$\phi \in \Form^+(Q,\vec{x})$, we define $\Post(\phi,u) \equiv \exists
\vec{x}_0 \ldots \exists \vec{x}_{n-1} ~.~
\Delta(\phi,u)[\vec{x}/\vec{x}_n]$.
